---
title: "Untitled"
output: html_document
date: "2024-06-19"
---

```{r load data}
library(tidyverse)
getwd()

#load("data/full_data.RData") # had an issue saving and loading the data!

attributes(full_data$datetime_posix) # already in correct TZ

subdata <- full_data %>% 
  filter(habitat == "slope") %>% 
  filter(! location %in% c("E_slope", "NW_slope")) %>% 
  #filter every fifth row
  filter(row_number() %% 5 == 1)  %>%
  #create day time factor
  mutate(hour = hour(datetime_posix),
         minutes = minute(datetime_posix), 
         time = hour + minutes/60) %>% 
  #select(-c(hour, minutes)) %>% 
  mutate(daynight = ifelse(time > 5 & time < 17, "day", "night"),
         daynight = factor(daynight)) %>% 
  select(HM, location, datetime_posix, period, month, daynight, day, hour, time, 
         low_freq_M, lunar_phase8, lunar_illum, temp, wd, ws) %>% 
    mutate(lunar_phase8 = factor(lunar_phase8, levels=c("New", "Waxing crescent", "First quarter", "Waxing gibbous", "Full", "Waning gibbous", "Last quarter", "Waning crescent"))) %>%
  mutate(lunar.n = as.numeric(lunar_phase8)) %>%  # model lunar phase as a continuous variable
  #Some extreme outlier maybe association with boat
  #Remove top quantiles for each group 
 group_by(location) %>% 
  group_modify(~ if (.y$location == "SW_slope") filter(.x, between(low_freq_M, quantile(low_freq_M, 0), quantile(low_freq_M, .99))) else filter(.x, between(low_freq_M, quantile(low_freq_M, 0), quantile(low_freq_M, .99)))) %>% 
  filter(between(low_freq_M, quantile(low_freq_M, 0), quantile(low_freq_M, .99))) %>% 
  mutate(stemp = scale(temp), 
         sws = scale(ws), 
         nLFM = (low_freq_M - min(low_freq_M))/(max(low_freq_M) - min(low_freq_M)))

subdata_lf <- subdata
```

Check distribution of M
```{r eval = FALSE}
subdata1 <- full_data %>% filter(habitat=="slope")

ggplot(subdata1, aes(x = datetime_posix, y = low_freq_M, col = location)) +
  geom_line()

ggplot(subdata1, aes(x = temp, fill = period)) +
  geom_histogram(bins=100)+
  facet_wrap(~location, scales = "free")
#there is an interaction between temp and time of year
ggplot(subdata1, aes(x = ws, fill = period)) +
  geom_histogram(bins=100)+
  facet_wrap(~location, scales = "free")
#there is an interaction between temp and time of year

ggplot(subdata1, aes(x = low_freq_M, fill = location)) +
  geom_histogram(bins=100)+
  facet_wrap(~location, scales = "free")

ggplot(subdata1, aes(x = low_freq_M, fill = location)) +
  geom_density()+
  facet_wrap(~location, scales = "free")
ggplot(subdata1, aes(x = low_freq_M, fill = location)) +
  geom_histogram(bins=100)+
  facet_wrap(~location, scales = "free")
# very long right tails
```


```{r eval = FALSE}
ggplot(subdata, aes(x = datetime_posix, y = low_freq_M, col = location)) +
  geom_line()

ggplot(subdata, aes(x = stemp, fill = period)) +
  geom_histogram()+
  facet_wrap(~location)
#there is an interaction between temp and time of year
ggplot(subdata, aes(x = sws, fill = period)) +
  geom_histogram()+
  facet_wrap(~location)
#there is an interaction between temp and time of year

ggplot(subdata, aes(x = low_freq_M, fill = location)) +
  geom_histogram(bins=100)+
  facet_wrap(~location)

ggplot(subdata, aes(x = low_freq_M, fill = location)) +
  geom_histogram()+
  facet_wrap(~location)
ggplot(subdata, aes(x = nLFM, fill = location)) +
  geom_histogram(bins=100)+
  facet_wrap(~location)
#Much better distributions... 
```

```{r}
subdata.test <- subdata %>% 
  filter(row_number() %% 9 == 1)  %>%
  filter(datetime_posix <= as.Date("2023-05-15")) 

subdata.partial <- subdata %>% 
  filter(row_number() %% 20 == 1)

subdata.full <- subdata 

ggplot() +
  geom_point(data = subdata.test, aes(y = nLFM, x = datetime_posix), col = "red", alpha = .2) +
  geom_point(data = subdata.partial, aes(y = nLFM, x = datetime_posix), col = "pink", alpha = .2) +
  geom_point(data = subdata.full, aes(y = nLFM, x = datetime_posix), col = "lightblue", alpha = .2) +
  facet_wrap(~location)

# how you subset seems to matter
```


# GAMs ----

```{r}
library(mgcv)      #for GAMs
library(gratia)    #for GAM plots
library(broom)     #for tidy output#
library(emmeans)   #for marginal means etc
library(MuMIn)     #for model selection and AICc
library(tidyverse) #for data wrangling
library(DHARMa)    #for simulated residuals
library(performance) #for residual disagnostics
library(see)        # to visualize residual diagnostics
library(ggpubr) # for combining final plots

```


# Fit the model

```{r}
gam_Lfreq1a <- gam(low_freq_M ~ s(time, bs = "cc", k = 40) +
                    s(location, bs = "re"), 
                  data = subdata.test, 
                  family = tw(),
                  method = 'REML')

gam_Lfreq1b <- gam(nLFM ~ s(time, bs = "cc", k = 40) +
                    s(location, bs = "re"), 
                  data = subdata.test, 
                  family = tw(),
                  method = 'REML')

simulateResiduals(gam_Lfreq1a,  plot=TRUE)
simulateResiduals(gam_Lfreq1b,  plot=TRUE)
#normalised data improves fit
```


## Fit model to partial data ----

```{r}
gam_Lfreq.partial <- gam(nLFM ~ s(time, bs = "cc", k = 40) +
                    s(lunar.n, bs = "cc", k = 5) +
                    #s(day, bs ="ps", k = 10) + #concurvity - soaks up variance from temp
                    s(stemp, bs ="cr", k = 10) +
                    s(sws, bs = "cr", k = 10) + #not important
                    s(location, daynight, bs = "re"), 
                  data = subdata.partial, 
                  family = tw(),
                  method = 'REML')
```


## Model validation
```{r}
k.check(gam_Lfreq.partial)
appraise(gam_Lfreq.partial)
concurvity(gam_Lfreq.partial)
resids <- simulateResiduals(gam_Lfreq.partial,  plot=TRUE)
testDispersion(resids)
```
Actually, quite a reasonable model

## Which distribution fits best?
```{r}
#Gamma <- gam(nLFM ~ s(time, bs = "cc", k = 40) +
#                    s(lunar_illum, bs = "cr", k = 10) +
#                    #s(day, bs ="ps", k = 10) + #concurvity - soaks up variance from temp
#                    s(stemp, bs ="cr", k = 10) +
#                    #s(sws, bs = "cr", k = 10) + #not important
#                    s(location, daynight, bs = "re"), 
#                  data = subdata.partial, 
#                  family = Gamma(),
#                  method = 'REML')

Beta <- gam(nLFM ~ s(time, bs = "cc", k = 40) +
                    s(lunar_illum, bs = "cr", k = 10) +
                    #s(day, bs ="ps", k = 10) + #concurvity - soaks up variance from temp
                    s(stemp, bs ="cr", k = 10) +
                    #s(sws, bs = "cr", k = 10) + #not important
                    s(location, daynight, bs = "re"), 
                  data = subdata.partial, 
                  family = betar(),
                  method = 'REML')

Tweedie <- gam(nLFM ~ s(time, bs = "cc", k = 40) +
                    s(lunar_illum, bs = "cr", k = 10) +
                    #s(day, bs ="ps", k = 10) + #concurvity - soaks up variance from temp
                    s(stemp, bs ="cr", k = 10) +
                    #s(sws, bs = "cr", k = 10) + #not important
                    s(location, daynight, bs = "re"), 
                  data = subdata.partial, 
                  family = tw(),
                  method = 'REML')

AIC(Beta, Tweedie) 
# Tweedie has highest AIC
# WHat about residuals?
#simulateResiduals(Gamma,  plot=TRUE)
simulateResiduals(Beta,  plot=TRUE)
simulateResiduals(Tweedie,  plot=TRUE)


```


## Fit model to full data ----

```{r}
gam_Lfreq.full <- gam(nLFM ~ s(time, bs = "cc", k = 40) +
                    #s(lunar_illum, bs = "cr", k = 10) +
                    s(lunar.n, bs = "cc", k = 5) +
                    #s(day, bs ="ps", k = 10) + #concurvity - soaks up variance from temp
                    s(stemp, bs ="cr", k = 10) +
                    s(location, daynight, bs = "re"), 
                  data = subdata, 
                  family = tw(),
                  method = 'REML')
```
# random effect for daynight?
```{r}
gam_loc <- gam(nLFM ~ s(time, bs = "cc", k = 40) +
                    s(lunar_illum, bs = "ts", k = 10) +
                    s(stemp, bs ="cr", k = 1) +
                    s(sws, bs = "cr", k = 1) + 
                    s(location, bs = "re"), 
                  data = subdata.full, 
                  family = betar(),
                  method = 'REML')

gam_loc_diel <- gam(nLFM ~ s(time, bs = "cc", k = 40) +
                    s(lunar_illum, bs = "ts", k = 10) +
                    s(stemp, bs ="cr", k = 1) +
                    s(sws, bs = "cr", k = 1) + 
                    s(location, daynight, bs = "re"), 
                  data = subdata.full, 
                  family = betar(),
                  method = 'REML')

AIC(gam_loc, gam_loc_diel) # adding daynight improves model

gam.check(gam_loc); gam.check(gam_loc_diel)

summary(gam_loc); summary(gam_loc_diel) # +1% dev. explained


```

# Full model:
```{r}
gam_Lfreq.full <- gam(nLFM ~ s(time, bs = "cc", k = 40) +
                    s(lunar.n, bs = "cc", k=5) +
                    #s(stemp, bs ="cr", k = 10) +
                    s(sws, bs = "cr", k = 10) + 
                    s(location, daynight, bs = "re"), 
                  data = subdata.full, 
                  family = tw(),
                  method = 'REML')
```

## Model validation
```{r}
k.check(gam_Lfreq.full)
appraise(gam_Lfreq.full)
concurvity(gam_Lfreq.full)
resids <- simulateResiduals(gam_Lfreq.full,  plot=TRUE)
testDispersion(resids)

# Not trimming quantile -> Overdispersion
# trimming quantile >95% -> Underdispersion
# trimming >99% -> dispersion is OK

```
Very reasonable

```{r}
my.gam = gam_Lfreq.full
```


# Partial plots
```{r}
draw(my.gam)
#draw(my.gam, residuals = TRUE)
```


# Model investigation / hypothesis testing {.tabset .tabset-faded}
```{r}
summary(my.gam)
```

Significant wiggliness throughout diurnal cycles, lunar cycles, and current changes

# Custom Plotting
``` {r}
sm <- smooth_estimates(my.gam) |>
  add_confint()

x <- subdata.full |>
  add_partial_residuals(my.gam)

names(x)

# Choose colours
library(RColorBrewer)
my_cols <- brewer.pal(n = 4, name = 'Dark2')
my_cols

# set font size:
custom_theme <- theme(
  axis.title = element_text(size = 5.5),
  axis.text = element_text(size = 5),)

# plotting time
p_time_LFM <- sm |>
  filter(.smooth == "s(time)") |>
  ggplot() +
  geom_ribbon(aes(ymin = .lower_ci, ymax = .upper_ci, x = time), 
              fill = my_cols[1], alpha = 0.2) +
  geom_line(aes(x = time, y = .estimate), colour = my_cols[1], lwd = 1) +  
      theme_bw() +
  labs(y = "M in the low freq. band (normalised)", x = "Hour of the Day")+
  scale_x_continuous(breaks = seq(0, 25, by = 2), expand = expansion(add=0))+
  scale_y_continuous(expand = expansion(add=0), limits = c(-2.1, 2.1),
                     labels = scales::number_format(accuracy = 0.1))+
  custom_theme +
  annotate("text", x = -Inf, y = Inf, label = "A", hjust = -0.5, vjust = 1.5, size = 3, fontface = "bold")

# Print the plot
  ggsave("../Plots/p_time_LFM.png", width=7, height=4)

# plotting lunar
p_lunar_LFM <- 
  sm |>
  filter(.smooth == "s(lunar.n)") |>
  ggplot() +
  geom_ribbon(aes(ymin = .lower_ci, ymax = .upper_ci, x = lunar.n),  
    fill = my_cols[3], alpha = 0.2) +
  geom_line(aes(x = lunar.n, y = .estimate), colour = my_cols[3], lwd = 1) + 
  labs(y = "Normalised Low freq. M")+
      theme_bw() +
  labs(y = "M in the low freq. band (normalised)", x = "Lunar Phase")+
  scale_x_continuous(breaks = seq(1, 8, by = 1), 
                     limits = c(1, 8), 
                     expand = expansion(add = 0),
                     labels = c("New", "", "Waxing", "", "Full", "", "Waning", "")) +
  scale_y_continuous(expand = expansion(add=0), limits = c(-0.51, 0.51))+
  custom_theme+
  annotate("text", x = -Inf, y = Inf, label = "B", hjust = -0.5, vjust = 1.5, size = 3, fontface = "bold")


# Print the plot
ggsave("../p_lunar_LFM.png", width=7, height=4)

# plotting current
p_current_LFM <- sm |>
  filter(.smooth == "s(sws)") |>
  ggplot() +
  geom_ribbon(aes(ymin = .lower_ci, ymax = .upper_ci, x = sws), 
    fill = my_cols[4], alpha = 0.2) +
  geom_line(aes(x = sws, y = .estimate), colour = my_cols[4], lwd = 1) + 
  theme_bw() +
  labs(y = "M in the low freq. band (normalised)", x = "Normalised Current Speed (m/s)")+
  scale_x_continuous(expand = expansion(add=0))+
  scale_y_continuous(expand = expansion(add=0), limits = c(-0.51, 0.51))+
  custom_theme +
  annotate("text", x = -Inf, y = Inf, label = "C", hjust = -0.5, vjust = 1.5, size = 3, fontface = "bold")


# Print the plot
ggsave("../Plots/p_current_LFM.png", width=7, height=4)


combo_plot <- 

  ggarrange(p_time_LFM, p_lunar_LFM, p_current_LFM, 
                           #labels = c("A", "B", "C"),
                           #hjust = -48, 
                           ncol = 1, nrow = 3,
                           align = "v",
                           heights = c(1, 1),
                           font.label = list(size = 14))

ggsave("../Plots/combo_plot_LFM.png", combo_plot)


```



